%%
%% PEARC'26 Extended Abstract
%% Practice & Experience in Advanced Research Computing 2026
%% July 26--30, 2026, Minneapolis, Minnesota, USA
%%
%% Authors: Geoffrey Lentner, Ashish
%% Affiliation: Rosen Center for Advanced Computing, Purdue University
%%

%% For initial submission, use [manuscript] per PEARC guidelines.
%% For camera-ready, switch to [sigconf].
\documentclass[manuscript]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information. These commands have SAMPLE values;
%% replace with values from your rights confirmation email after acceptance.
\setcopyright{acmcopyright}
\copyrightyear{2026}
\acmYear{2026}
\acmDOI{XXXXXXX.XXXXXXX}  % TODO: Update after acceptance

%% Conference metadata for PEARC'26
\acmConference[PEARC'26]{Practice and Experience in Advanced Research Computing 2026}{July 26--30, 2026}{Minneapolis, MN, USA}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{PEARC'26: Practice and Experience in Advanced Research Computing 2026,
%%  July 26--30, 2026, Minneapolis, MN, USA}
\acmISBN{978-1-4503-XXXX-X/2026/07}  % TODO: Update after acceptance

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

\begin{document}

\title{Hello Computer: HPC in the Agentic Era}

%%
%% Authors
%%
\author{Geoffrey Lentner}
\email{glentner@purdue.edu}
\orcid{0000-0001-9314-0683}
\affiliation{%
  \institution{Rosen Center for Advanced Computing}
  \city{West Lafayette}
  \state{Indiana}
  \country{USA}
}

\author{Ashish}
\email{ashish@purdue.edu}
\orcid{0009-0008-1243-1306}
\affiliation{%
  \institution{Rosen Center for Advanced Computing}
  \city{West Lafayette}
  \state{Indiana}
  \country{USA}
}

\renewcommand{\shortauthors}{Lentner and Ashish}

\begin{abstract}
Large language models have evolved from curiosity to co-pilot in under four years.
With the emergence of agentic AI systems that reason, plan, and execute multi-step
tasks autonomously, HPC centers face a new category of user need: researchers expect
these tools to be supported, not just permitted. This paper offers
a practitioner's perspective from Purdue's Rosen Center for Advanced Computing (RCAC),
where we have begun deploying system-wide configurations, custom MCP servers, and user
guidance for agentic tools. As a demonstration, every word of this manuscript was
produced through an agent-first workflow: over sixty commits of iterative collaboration
between human authors and AI agents, documented in a public GitHub repository. We argue
that proactive engagement, not prohibition, is the path forward for facilitators who
wish to remain relevant in the agentic era.
\end{abstract}

%%
%% TODO: Generate CCS codes at https://dl.acm.org/ccs/ccs.cfm
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
%% \begin{CCSXML}
%% <ccs2012>
%%  <concept>
%%   <concept_id>00000000.0000000.0000000</concept_id>
%%   <concept_desc>Replace with actual CCS concepts</concept_desc>
%%   <concept_significance>500</concept_significance>
%%  </concept>
%% </ccs2012>
%% \end{CCSXML}
%%
%% \ccsdesc[500]{Replace with actual CCS concepts}

\keywords{high-performance computing, agentic AI, large language models, Model Context Protocol, context engineering, research computing facilitation}

\maketitle

\section{Shall We Play a Game?}

On November 30, 2022, OpenAI released ChatGPT as a ``research preview.''
Within five days it had one million users; by January 2023, one hundred
million, the fastest-growing consumer application in
history~\cite{openai2022chatgpt}. The inflection point was not the technology itself
but its sudden public accessibility: for the first time, anyone could converse with a
frontier language model.
For those of us supporting researchers on high-performance computing systems, the
implications arrived slowly and then all at once. At first, ChatGPT was a curiosity, a
toy that could write passable Python functions or explain error messages with
varying degrees of accuracy. Then came GPT-4~\cite{openai2023gpt4},
Claude~\cite{anthropic2024claude3}, and their competitors, each iteration more capable
than the last. Models became multimodal, processing images, audio, and video alongside
text. Then came \emph{agentic} AI: systems that don't just respond to prompts but
reason, plan, and execute multi-step workflows autonomously~\cite{yao2023react}.
Today's agentic development environments (Warp, Cursor, Claude Code) can read files,
run shell commands, edit code, and iterate toward solutions with increasing
sophistication. HPC centers now face a question with no precedent in our field: how do
we engage with tools that can, in some cases, do our jobs faster than we can?

%% \subsection{Mostly Harmless}

\emph{The Hitchhiker's Guide to the Galaxy} famously summarized Earth in two words:
``Mostly Harmless.'' We adopt this as our working hypothesis for agentic AI in HPC.
Our assessment at Purdue's Rosen Center for Advanced Computing (RCAC) is that these
tools are, in fact, mostly harmless. Frontier AI systems follow instructions more
reliably than many users do: they respect queue policies, observe resource limits, and
avoid dangerous operations when properly configured. The greater risk lies in
disengagement. A graduate student debugging a CUDA kernel at 2~AM will reach for ChatGPT
whether or not we endorse it. If we don't shape the context in which that interaction
occurs, we leave both the student and the cluster exposed to preventable mistakes.
We chose proactive engagement. We use these tools ourselves, daily, so that we have
visceral understanding of their capabilities and limitations. This paper is itself a
product of that practice: an agent-first workflow comprising rules files, session logs,
annotated references, and iterative revision, all documented in a public GitHub
repository. This paper documents our early experience deploying system-wide
configurations, custom MCP (Model Context Protocol) servers, and user guidance. We
offer this as both evidence and invitation.

\section{Background}

The trajectory from research curiosity to practical tool was swift. In 2017, Vaswani et
al.~introduced the Transformer architecture, replacing recurrence with attention
mechanisms~\cite{vaswani2017attention}. By 2020, GPT-3's 175 billion parameters
demonstrated that scale alone could yield emergent capabilities~\cite{brown2020gpt3}.
ChatGPT's November 2022 launch brought these capabilities to the general public,
reaching 100 million users in two months~\cite{openai2022chatgpt}.

A parallel development transformed chatbots into \emph{agents}. The ReAct framework
formalized this shift: agentic systems interleave reasoning traces with task-specific
actions, interfacing with external tools and knowledge sources rather than relying solely
on training data~\cite{yao2023react}. Today's agentic development environments implement
this pattern: reading files, executing commands, editing code, and iterating based on
observed results~\cite{warp2024agentmode}.

The infrastructure for agentic AI has matured rapidly. Anthropic's Model Context Protocol
(MCP), released November 2024, provides a standardized interface for AI systems to connect
with external data sources and tools, analogous to how the Language Server Protocol
unified IDE tooling~\cite{anthropic2024mcp}. By March 2025, OpenAI and Google had adopted
MCP~\cite{openai2025mcp}; by December, Anthropic donated it to the Linux Foundation's
Agentic AI Foundation, with over 10,000 active public servers~\cite{anthropic2025aaif}.

For HPC, this convergence is consequential. Deelman et al.\ argue that high-performance
computing stands ``at a crossroads''~\cite{deelman2025hpc}. The Genesis Mission executive
order (November 2025) charges DOE with uniting national laboratories, supercomputers, and
AI systems for scientific discovery~\cite{whitehouse2025genesis}. The question is no
longer whether AI will integrate with HPC workflows, but how, and how centers choose to
engage will define the experience for researchers and facilitators alike.

\section{Approach}

Our approach at RCAC combines three complementary strategies: system-wide agent
configurations, purpose-built MCP servers, and user documentation that acknowledges
the reality of agentic tool adoption.

\paragraph{System-Wide Configurations}

Modern agentic tools look for rules files in well-known
locations. We deploy cluster-wide configurations through an \texttt{/etc/agents.d}
directory hierarchy, injecting HPC-specific context: which commands to use for quota
checks (\texttt{myquota}) and batch job submission (\texttt{sbatch}), how to load
software via environment modules, and which filesystems serve which purposes. These
configurations also encode prohibitions: don't run computationally intensive work on
login nodes, don't store sensitive data in world-readable locations, don't submit jobs
without time limits. The agent absorbs the cluster's policies before the user asks
their first question.

\paragraph{MCP Servers}

We developed two purpose-built MCP servers (the Model Context Protocol, not the Master
Control Program, though the naming coincidence feels appropriate): RCAC-MCP, which
exposes Slurm operations, filesystem navigation, and cluster-specific tools like
\texttt{myquota} and \texttt{jobinfo}; and Globus-MCP, which provides agentic data
transfers between storage endpoints. Both are publicly available on GitHub.
Both servers implement a local-first architecture. The MCP server runs as
a subprocess of the user's IDE, communicating via stdin/stdout. Commands execute remotely 
through the user's existing SSH configuration: no new credentials, no hosted infrastructure,
no additional attack surface. If you can SSH to the cluster, your agent can too. For
environments requiring centralized control, we also support hosted deployment with
JWT/OIDC authentication and identity delegation.

\paragraph{Documentation and Guidance}

Perhaps most importantly, we updated our user documentation to address agentic tools
directly. We don't pretend they don't exist or discourage their use. We explain what
they can and cannot do, how to verify their suggestions, and how to report issues when
AI-generated commands go wrong. We treat agentic AI as we would any other tool in the
HPC ecosystem: powerful, useful, and requiring appropriate guidance.

\section{Discussion}

\paragraph{``I'm Sorry, Dave'' (User Support)}

The support dynamic is shifting. Users increasingly arrive at our help desk having already
consulted an AI, sometimes with correct solutions, sometimes with confident hallucinations.
We've seen tickets where the user's first message includes an AI-generated Slurm script
that would work perfectly... on a cluster we don't operate.
The irony isn't lost on us. When HAL refused to open the pod bay doors, the problem was
an AI saying no. Our challenge is different: AI that says yes too readily, without
understanding the constraints of a particular environment. The solution isn't to compete
with AI for first response; it's to ensure the AI has the context to respond correctly.
Hence our investment in system prompts and MCP servers that know our clusters.

\paragraph{``Tea, Earl Grey, Hot'' (Context Engineering)}

Captain Picard never asked the replicator for ``a beverage.'' He specified exactly what
he wanted: ``Tea, Earl Grey, hot.'' Effective use of agentic AI demands similar precision,
and the best results come from engineering that precision into the system alongside the
user, not expecting either to carry the burden alone.
Context engineering is the practice of structuring information, rules, and reference
material so that an agent produces reliable output from the outset. We developed this
discipline through a year of iterative practice before beginning this paper: learning
which context formats yield consistent results, how to manage context window capacity
across long workflows, and how to decompose complex tasks into verifiable steps. The
same principle applies to HPC. ``Submit a job'' fails where ``submit a GPU job to the
Gautschi cluster using my lab's allocation'' succeeds. System-wide configurations
provide the foundation, but users still bring the domain specificity that makes each
interaction productive.

\paragraph{``I Know Kung Fu'' (User Expectations)}

Neo's instant kung fu download is a fantasy that dies hard. Users sometimes expect AI
to make them instant experts, to transform a novice into an HPC specialist through
conversation alone. The reality is more nuanced: AI makes you faster at what you already
understand, not expert at what you don't.
This creates what we call the expertise paradox: you need domain knowledge to verify
whether AI output is correct, but the people most likely to rely heavily on AI are
those without that knowledge. Our documentation explicitly addresses this: use AI to
accelerate your learning, not to bypass it. Ask it to explain \emph{why}, not just
\emph{what}.
The goal is augmented expertise, not outsourced understanding.

%% \paragraph{``The Answer is 42'' (AI Outcomes)}
%% 
%% Deep Thought computed the answer to Life, the Universe, and Everything: 42. The answer
%% was correct. The question was wrong. We've experienced both extremes: an agent that
%% built our thirteen-source annotated bibliography across fifteen context windows over
%% eight hours of deliberate prompting, and the same agent confidently generating Slurm
%% commands for schedulers we don't use. Plausible output
%% requiring expert verification is consistent with recent evaluations of LLMs for HPC
%% software development~\cite{godoy2024llm}. Verification isn't optional; it's the core
%% competency that separates productive use from cargo cult computing. The answer may be
%% 42, but only if you asked the right question.

\paragraph{``Don't Cross the Streams'' (Cautionary Notes)}

Egon's warning about crossing the streams applies to agentic AI in HPC: some combinations
are dangerous. An agent with shell access can \texttt{rm -rf} a project directory. An
agent managing job submissions can exhaust an allocation in hours. An agent reading
environment variables might log API keys to a context window that persists beyond the
session.
But these risks are not new. Users have always been capable of destroying their own data,
exhausting their allocations, and exposing credentials. Mature HPC centers already build
confinement: cgroups to isolate resources, node health checks, root-squashed filesystems,
quota enforcement. These mechanisms contain the blast radius so that one user's mistake
cannot cascade across the system. Most agentic tools also support user confirmation for
destructive actions out of the box, providing an additional safety net. Agents accelerate
the pace of potential errors, which
means existing hardening must be strengthened, not invented from scratch. We are also
developing practices for capturing agent interactions that contribute to scientific
outcomes, because reproducibility in AI-assisted research remains an open question.

\section{End of Line}

As a meta-demonstration, this paper was itself produced through an agent-first workflow.
No prose was written by hand. Human authors defined rules, curated references, reviewed
transcripts, and guided iterative revision; AI agents executed the research, planning,
drafting, and integration. The entire process, comprising over sixty commits, session
logs, and planning documents, is available in the accompanying GitHub
repository.\footnote{\url{https://github.com/glentner/pearc26-hello-computer}} The
methodology itself may be a contribution worth exploring.

We began this paper with a question: how should HPC centers engage with tools that can, in
some cases, outperform us at our own jobs? Our answer has been proactive engagement, not
prohibition.
At RCAC, we're building the infrastructure to make agentic AI a first-class citizen of the
HPC ecosystem. 
%% :system-wide configurations that expose cluster metadata, MCP servers that
%% provide contextual help, and documentation that prepares users for both the capabilities
%% and limitations of these tools. 
This work is early. We don't yet know what patterns will
emerge, what pitfalls await, or how dramatically these tools will reshape our profession.
The agentic era has arrived. HPC facilitators who engage early will shape its trajectory;
those who wait may find themselves shaped by it. We choose engagement, and we invite the
community to join us.
\emph{End of line.}

%%
%% Acknowledgments
%%
\begin{acks}
\textbf{Use of Generative AI.}
In accordance with ACM policy, we disclose that this manuscript was produced through an agent-first workflow. 
%% using Claude 4.6 Opus (Anthropic) via the Warp agentic development environment. 
AI agents performed research, drafting, revision, and project organization under human direction. 
All content was reviewed, verified, and approved by the human authors, who take full responsibility for the work.
%% This work was supported by the Rosen Center for Advanced Computing (RCAC) at Purdue University.
\end{acks}

%%
%% Bibliography
%%
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
